---
title: "True_Final"
author: "Phil Kaplan"
date: "2023-12-22"
output: html_document
bibliography: Final_Bibliography.bib
---
```{r setup, echo=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(MuMIn)
library(effects)
library(car)
library(broom)
library(knitr)
library(kableExtra)

colony_data <- read.csv("Colony.csv")
stressor_data <- read.csv("Stressor.csv")
```



``` {r data prep, echo=FALSE, warning=FALSE}
# Function to expand rows with month ranges
expand_months <- function(row, toggle) {
  year <- row$year
  state <- row$state

  # Extract start and end months
  months_range <- strsplit(as.character(row$months), "-")[[1]]
  start_month <- match(months_range[1], month.name)
  end_month <- match(months_range[length(months_range)], month.name)

  # Generate a sequence of months
  expanded_months <- seq(as.Date(paste(year, start_month, "01", sep = "-")),
                         as.Date(paste(year, end_month, "01", sep = "-")),
                         by = "1 month")

  # Create a data frame with expanded rows for stressor
  if (toggle == 0) {
    data.frame(
      year = row$year,
      month = format(expanded_months, "%B"),
      state = rep(state, length(expanded_months)),
      stressor = row$stressor,
      stress_pct = row$stress_pct
    )
    # Create a data frame with expanded rows for colony
  } else {
    data.frame(
      year = row$year,
      month = format(expanded_months, "%B"),
      state = rep(state, length(expanded_months)),
      colony_n = row$colony_n,
      colony_max = row$colony_max,
      colony_lost = row$colony_lost,
      colony_lost_pct = row$colony_lost_pct,
      colony_added = row$colony_added,
      colony_reno = row$colony_reno,
      colony_reno_pct = row$colony_reno_pct
    )
  }
}

# Apply the function to each row of stressors
expanded_stressor_data <- stressor_data %>%
  rowwise() %>%
  do(expand_months(.,0))

# Combine the expanded data with the original data
new_stressor_data <- bind_rows(expanded_stressor_data)

expanded_colony_data <- colony_data %>%
  rowwise() %>%
  do(expand_months(.,1))

# Combine the expanded data with the original data
new_colony_data <- bind_rows(expanded_colony_data)

# Combine colony and stressor data
merged_data_with_total <- left_join(new_colony_data, new_stressor_data, by = c("year", "month", "state"))

# Get rid of total value
merged_data <- merged_data_with_total %>% 
  filter(state != "United States") %>%
  mutate(stressor = ifelse(stressor == "Disesases", "Diseases", stressor)) %>%
  distinct() %>%
  na.omit()

# Columns we will use for analysis
selected_columns <- merged_data %>%
  select(state, year, month, stressor, colony_lost, colony_lost_pct, stress_pct)
```

# Introduction

Bees are crucial to humans’ everyday life, holding cultural and environmental importance. As pollinators and producers of honey and medicinal products, bees control the growth and quality of vegetation. The movement of pollen between plants is necessary for plants to fertilize and reproduce. However, there’s been evidence of bee colony loss over the years due to many factors including parasites, pesticides, lack of flowers, diseases, pests, mites, and even weather. These stressors have been found to accumulate into stress for the bees. During winter 2022-2023, an estimated 37.4% of managed colonies in the United States were lost. This winter loss rate was 13.2 percentage point (pp) in excess of the previous winter loss rate (24.2%), and 9.1 pp higher than the average winter loss (28.2%, 15-year average) reported by beekeepers since the start of the survey in 2008, making 2022-2023 the second highest year of winter loss after 2018-2019 @Steinhauer2023. It’s been observed and reported that chronic exposure to multiple interacting stressors is driving honey bee colony losses and declines of wild pollinators, but the precise combination apparently differs from place to place. For example, some pesticides act synergistically more than additively. Both pesticide exposure and food stress can decrease immune responses, making bees more susceptible to parasites @Goulson2015. In this experiment, we look at data that includes colony loss, what type of stressor, state, and even month to determine the correlation and relationships between these factors to bee colony death.



# Methods

Our research focused on bee colonies in terms of the number of colonies, maximum, lost, percent lost, added, renovated and percent renovated, as well as the number of colonies lost with Colony Collapse Disorder symptoms with both over and less than five colonies, also identifying colony health stressors with fiver or more colonies, leveraging data from honey bee colonies from the USDA, with stressors including state, varroa mites, other pests or parasites, diseases, pesticides, and other or unknown factors. This data was pooled from all across the United States, including Alabama, Arizona, Arkansas, California, Colorado, Connecticut, Florida, Georgia, Hawaii, Idaho, Illinois, Indiana, Iowa, Kansas, Kentucky, Louisiana, Maine, Maryland, Massachusetts, Michigan, Minnesota, Mississippi, Missouri, Montana, Nebraska, New Jersey, New Mexico, New York, North Carolina, North Dakota, Ohio, Oklahoma, Oregon, Pennsylvania, South Carolina, South Dakota, Tennessee, Texas, Utah, Vermont, Virginia, Washington, West Virginia, Wisconscin, Wyoming, and other states, with data spanning from 2015 to 2021.  



``` {r methods graphs, echo=FALSE, warning=FALSE, message=FALSE}
# Visualize relationship between Stress Percentage and Colony Loss Percentage
ggplot(merged_data, aes(x = stress_pct, y = colony_lost_pct)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm", color = "red") +
  labs(title = "Relationship between Stress Percentage and Colony Loss",
       x = "Stress Percentage",
       y = "Colony Loss Percentage") + 
  theme_minimal()

# Get mean colony lost percentage
mean_colony_lost_pct_tibble <- selected_columns %>%
  group_by(state) %>%
  summarize(mean_colony_lost_pct = mean(colony_lost_pct))

# Graph of Colony Losses by State
ggplot(mean_colony_lost_pct_tibble, aes(x = reorder(state, -mean_colony_lost_pct), y = mean_colony_lost_pct, fill = state)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank()) +
  labs(title = "Colony Loss Percentage by State",
       x = "State",
       y = "Colony Loss Percentage")

# Graph function for column correlation w/ Colony Loss percentage
create_correlation_plot <- function(x_variable) {
  # Calculate the mean and standard deviation within the ggplot code
  plot <- ggplot(selected_columns, aes_string(x = x_variable, y = "colony_lost_pct")) +
    stat_summary(fun.data = "mean_se", geom = "point", position = position_dodge(width = 0.5)) +
    stat_summary(fun.data = "mean_se", geom = "errorbar", position = position_dodge(width = 0.5)) +
    labs(title = paste("Correlation between", x_variable, "and Colony Lost Percentage"),
         x = x_variable,
         y = "Colony Lost Percentage") +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  return(plot)
}

create_correlation_plot("stressor")

month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Convert "month" to a factor with the custom order
selected_columns$month <- factor(selected_columns$month, levels = month_order)

create_correlation_plot("month")
```



# Results

``` {r models, echo=FALSE,warning=FALSE, message=FALSE }
fit1 <- lm(colony_lost_pct ~ stress_pct *stressor*month+state, data = selected_columns,na.action = "na.fail")
fit_dr <- dredge(fit1)
fit_dr %>% view
fit_best <- get.models(fit_dr,subset = 1)[[1]]
Anova(fit_best)
```

# Discussion

  The ANOVA test of the best fitting model indicates that variables month, state, stress_pct, and stressor and interactions month:stress_pct, stress_pct:stressor, and month:stress_pct:stressor are statistically significant predictors of colony loss percentage. Notably, interaction month:stressor contributes to the best model but is not statistically significant.
  The graph of colony loss percentage versus stress percentage indicates that there is a positive relationship between these two variables. The graph of colony loss percentage by state indiates that Alabama, Arizona, and Arkansas had the greatest colony percentage. There is not a clear relationship between relative geographic region and colony loss percentage from the graph although state is a statistically significant predictor. The graph of colony loss percentage by stressor indicates that pesticides and diseases are the greatest known contributers to the percent of colony loss. This is consistent with the findings of @Goulson2015 which indicated that pesticide exposure can decrease bee immune responses. The graph of colony loss percentage vs state indicates that the months with the greatest percentage of colony loss occur between October - March. This is consistent with the findings of @Steinhauer2023 that the greatest colony loss often occurs during the winter months. Notably, conditions over the summer are a significant predictor of winter survival @calovi2021summer. 
  
# Contributions
Stacey Kim: Sources, Introduction and Analysis
Tara Gerjarusak: Analysis and Methods
Blake Prall: Data Prep, Graphs, and Analysis
Phil: Code & Discussion 

# Sources





